%% MoCafe v2.00 -- Physics and Algorithms
%% Author: Kwang-Il Seon (KASI)
\documentclass[english,a4paper]{kiseon_memo}
\synctex=1

\usepackage{booktabs}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{microtype}

\newcommand{\kabs}{\kappa_{\mathrm{abs}}}
\newcommand{\kext}{\kappa_{\mathrm{ext}}}
\newcommand{\ksca}{\kappa_{\mathrm{sca}}}
\newcommand{\Cabs}{C_{\mathrm{abs}}}
\newcommand{\Cext}{C_{\mathrm{ext}}}
\newcommand{\Csca}{C_{\mathrm{sca}}}
\newcommand{\Jlam}{J_\lambda}
\newcommand{\khat}{\hat{\mathbf k}}
\newcommand{\nhat}{\hat{\mathbf n}}
\newcommand{\rvec}{\mathbf r}
\newcommand{\dd}{\mathrm{d}}
\newcommand{\code}[1]{\texttt{#1}}
\newcommand{\file}[1]{\texttt{#1}}
\newcommand{\pp}[1]{\texttt{par\%#1}}

\begin{document}

\title{MoCafe v2.00 --- Physics and Algorithms}
\author{Kwang-Il Seon (KASI)}
\date{2026-07-11}
\maketitle

\begin{abstract}
\noindent
This note documents the physics and the numerical algorithms behind MoCafe, the
parallel Monte-Carlo dust radiative-transfer code.  It develops the radiative
transfer problem and the Monte-Carlo solution (photon packets, free-path
sampling, forced first scattering, weighting, and Russian roulette), the
peeling-off (next-event) estimator, the scattering physics (Henyey--Greenstein
and the polarized Mueller-matrix path), panchromatic transport by grey
rescaling, the Lucy~(1999) path-length estimator for the radiation field, the
two dust-emission solvers (the Lucy iteration coupled to the \textsc{SEDust}
grain library with quantum-mechanical stochastic heating, and the
Bjorkman \& Wood immediate-reemission scheme), the Modified Random Walk, the
galaxy source models, the random-sampling routines (inversion, rejection, and
Walker's alias method), and the HEALPix interior observer.  Derivations are
given at the level needed to read and modify the code; the companion user guide
covers installation, the input reference, and the outputs.
\end{abstract}

\tableofcontents
\clearpage

%=============================================================================
\section{The radiative transfer problem}
%=============================================================================

\subsection{The transfer equation}
The specific intensity $I_\lambda(\rvec,\nhat)$ of radiation of wavelength
$\lambda$ at position $\rvec$ travelling in direction $\nhat$ obeys the
radiative transfer equation
\begin{equation}
  \nhat\!\cdot\!\nabla I_\lambda
   = -\kext(\rvec,\lambda)\,I_\lambda
     + \kext(\rvec,\lambda)\,S_\lambda(\rvec,\nhat),
  \label{eq:rte}
\end{equation}
where $\kext=\ksca+\kabs$ is the extinction coefficient (per unit length),
$\ksca$ and $\kabs$ the scattering and absorption coefficients, and $S_\lambda$
the source function.  For a dusty medium the source function has a scattering
part and a thermal-emission part,
\begin{equation}
  S_\lambda(\rvec,\nhat)
   = \frac{\ksca}{\kext}\!\int\! \frac{\dd\Omega'}{4\pi}\,
       \Phi(\nhat,\nhat')\,I_\lambda(\rvec,\nhat')
   + \frac{\kabs}{\kext}\, B_\lambda\!\big(T_{\rm d}(\rvec)\big),
  \label{eq:source}
\end{equation}
with $\Phi$ the (normalized) scattering phase function, $a\equiv\ksca/\kext$ the
single-scattering albedo, and $B_\lambda(T_{\rm d})$ the thermal emission of dust
at temperature $T_{\rm d}$.  Defining the optical depth along a ray by
$\dd\tau_\lambda=\kext\,\dd\ell$, the formal solution of Eq.~\eqref{eq:rte} is
\begin{equation}
  I_\lambda(\tau_\lambda)= I_\lambda(0)\,e^{-\tau_\lambda}
     + \int_0^{\tau_\lambda}\!\! S_\lambda(\tau')\,e^{-(\tau_\lambda-\tau')}\,\dd\tau',
  \label{eq:formal}
\end{equation}
i.e.\ attenuated incident light plus the attenuated contribution of every source
point along the ray.

\subsection{Why Monte Carlo}
Equations~\eqref{eq:rte}--\eqref{eq:source} couple every direction and position
(scattering redistributes light in angle; thermal emission couples wavelengths
through the temperature).  Rather than discretizing the integro-differential
operator, MoCafe evaluates the physical content of Eq.~\eqref{eq:formal}
\emph{statistically}: it follows a large number of photon packets whose random
histories --- emission, propagation over an exponentially distributed optical
depth, scattering into a new direction, absorption --- reproduce the transfer
equation in expectation.  Integrals over the phase function, the size
distribution, and the wavelength grid become sums over sampled events, so the
cost scales with the number of packets rather than with the dimensionality of
the problem, and complex three-dimensional geometries (Cartesian, clumpy, AMR)
are handled by the same estimator.

%=============================================================================
\section{Monte-Carlo photon transport}
%=============================================================================

\subsection{Photon packets and weights}
MoCafe transports photon \emph{packets}: weighted bundles of monochromatic
photons.  A packet carries a position $\rvec$, a propagation direction $\khat$,
a dimensionless weight $w$, a wavelength bin $\lambda$, a Stokes vector
$(I,Q,U,V)$ when polarization is on, and---in SED/emission mode---a physical
luminosity $L_{\rm p}$ [erg\,s$^{-1}$] so that energy budgets are absolute.  The
weight absorbs all the variance-reduction bookkeeping (forced scattering,
albedo, scan reweighting); the packet contributes $w\,L_{\rm p}$ wherever it is
tallied.

\subsection{Sampling the free path}
The probability that a packet reaches optical depth $\tau$ without interacting is
$e^{-\tau}$, so the interaction optical depth has probability density
$p(\tau)=e^{-\tau}$ and cumulative $P(\tau)=1-e^{-\tau}$.  Inverting
$P(\tau)=\xi$ for a uniform deviate $\xi\in[0,1)$ gives the standard free-path
sample
\begin{equation}
  \tau = -\ln(1-\xi).
  \label{eq:freepath}
\end{equation}
The sampled $\tau$ is converted to a physical displacement by integrating
$\kext$ along the ray until the accumulated optical depth equals $\tau$.  The
three media do this with a shared interface: Amanatides--Woo voxel traversal on
the Cartesian grid, a ray--sphere intersection followed by a DDA walk over the
clump acceleration grid in the clumpy medium, and an octree neighbour walk on the
AMR grid.  Because scattering and peeling-off call these tracers through a
procedure pointer, they are written once and are grid-agnostic.

\subsection{Forced first scattering and reduced weight}
In an optically thin medium most packets would escape without interacting,
which wastes samples.  Setting \pp{use\_reduced\_wgt} forces the first
interaction to occur inside the grid.  Let $\tau_0$ be the total optical depth
from the birth point to the far edge along $\khat$.  Restricting the free-path
distribution to $[0,\tau_0]$ and renormalizing gives the truncated density
$p(\tau)=e^{-\tau}/(1-e^{-\tau_0})$, whose inverse is
\begin{equation}
  \tau = -\ln\!\big[\,1-\xi\,(1-e^{-\tau_0})\big],\qquad \tau\in[0,\tau_0].
\end{equation}
To keep the estimator unbiased the packet weight is multiplied by the
probability that it would have interacted at all,
\begin{equation}
  w \leftarrow w\,(1-e^{-\tau_0}),
\end{equation}
and the complementary fraction $e^{-\tau_0}$ --- the probability of escaping
without scattering --- is carried by the direct (unscattered) peel-off toward
each observer, so no energy is lost.  This is why the $n_{\rm scatt}=0$ step is
special-cased in the transport loop.

\subsection{Absorption, weighting, and Russian roulette}
At an interaction MoCafe uses \emph{albedo weighting} rather than a stochastic
absorb/scatter decision: the packet always scatters, and the absorbed fraction
is accounted continuously (the scattered peel carries a factor $a$; the absorbed
energy is deposited into the radiation-field tally, \S\ref{sec:jlam}).  This
removes the variance of the absorb/scatter coin.  A packet whose weight has
decayed below a cutoff is subjected to Russian roulette: with survival
probability $q$ it continues with weight $w/q$, otherwise it is terminated.
Roulette keeps the estimator unbiased while preventing unbounded tracking of
negligible packets, and is used both in the transport loop and inside the
Modified Random Walk (\S\ref{sec:mrw}).

%=============================================================================
\section{The peeling-off (next-event) estimator}
%=============================================================================

\subsection{Derivation}
Waiting for packets to reach a distant observer by chance is hopelessly
inefficient.  Instead, at every emission and scattering site MoCafe adds the
\emph{expected} contribution the packet would make toward each observer --- the
peeling-off or next-event estimator.  Consider a scattering at position $\rvec$
with incoming direction $\khat$.  The joint probability that the packet scatters
into the small solid angle $\dd\Omega_o$ about the observer direction $\nhat_o$
\emph{and} then escapes to the observer without further interaction is
\begin{equation}
  a\,\frac{\Phi(\khat\!\cdot\!\nhat_o)}{4\pi}\,\dd\Omega_o \times e^{-\tau_o},
\end{equation}
where $a$ is the albedo, $\Phi$ the normalized phase function, and $\tau_o$ the
optical depth from $\rvec$ to the observer along $\nhat_o$.  For an observer of
collecting area $\dd A$ at distance $d_o$, $\dd\Omega_o=\dd A/d_o^2$, so the
image-plane contribution per unit area is
\begin{equation}
  \Delta I_o = \frac{w\,a\,\Phi(\khat\!\cdot\!\nhat_o)}{4\pi}\,
               \frac{e^{-\tau_o}}{d_o^{\,2}}.
  \label{eq:peel}
\end{equation}
The $1/d_o^2$ is the geometric dilution, $e^{-\tau_o}$ the attenuation to the
observer (computed by ray-tracing to the grid edge), and the phase-function
factor the probability of redirection.  Summing Eq.~\eqref{eq:peel} over all
scattering events of all packets builds the scattered image with far lower
variance than collecting escaping packets, because \emph{every} event
contributes to \emph{every} observer.

\subsection{Direct and scattered peels}
The unscattered (direct) light is peeled separately: a source packet contributes
its birth luminosity attenuated by the optical depth from the source to the
observer, forming the \code{Direct} image (and \code{Direct0}, the unattenuated
version, when requested).  Subsequent scatterings feed the \code{Scattered}
image through Eq.~\eqref{eq:peel}.  For an internal all-sky observer
(\S\ref{sec:allsky}) the same expected contribution is binned into the HEALPix
pixel of the arrival direction instead of a tangent-plane pixel.  All peels go
through the \code{raytrace\_to\_edge} procedure pointer, so the Cartesian,
clumpy, and AMR media share identical scattering and peeling code; the default
binding keeps the Cartesian output bit-identical across refactors.

%=============================================================================
\section{Scattering}
%=============================================================================

\subsection{Henyey--Greenstein}
The default phase function is Henyey--Greenstein with asymmetry parameter
$g=\langle\cos\theta\rangle$,
\begin{equation}
  \Phi_{\rm HG}(\mu) = \frac{1-g^2}{(1+g^2-2g\mu)^{3/2}},\qquad \mu=\cos\theta,
\end{equation}
normalized so that $\tfrac12\!\int_{-1}^{1}\Phi_{\rm HG}\,\dd\mu = 1$.  Its
cumulative distribution inverts in closed form, so the scattering cosine is
sampled directly from a uniform deviate $\xi$:
\begin{equation}
  \mu = \frac{1}{2g}\left[1+g^2-\left(\frac{1-g^2}{1-g+2g\xi}\right)^{2}\right]
        \quad (g\neq0),
\end{equation}
with the isotropic limit $\mu=2\xi-1$ for $g=0$.  The azimuth is uniform, and
the new direction is built by rotating $\khat$ by $(\theta,\varphi)$.

\subsection{Polarization: Stokes vectors and the Mueller matrix}
When a scattering-matrix table is supplied (\pp{scatt\_mat\_file}) the code
carries the Stokes vector $\mathbf S=(I,Q,U,V)^{\!\top}$ and replaces the scalar
phase function by the $4\times4$ Mueller matrix $\mathbf M(\theta)$ of the
grains.  A scattering event proceeds in three steps: (i) rotate $\mathbf S$ into
the scattering plane by the Stokes rotation $\mathbf R(\psi_1)$; (ii) apply the
Mueller matrix, $\mathbf S'=\mathbf M(\theta)\,\mathbf R(\psi_1)\,\mathbf S$;
(iii) rotate the result into the new meridian frame by $\mathbf R(\psi_2)$.  The
scattering angle is drawn from the tabulated $S_{11}(\theta)$ (the intensity
phase function) with an alias table (\S\ref{sec:alias}) for $\mathcal O(1)$
sampling; because $S_{11}$ already encodes the angular dependence, the azimuth is
drawn from the $S_{12}$-modulated distribution to respect the incident
polarization.  The peel-off of Eq.~\eqref{eq:peel} then uses the
matrix-weighted phase function evaluated for the observer direction, so the
polarized images ($Q,U,V$ planes) are next-event estimates just like $I$.  With
Henyey--Greenstein, \pp{use\_stokes} is forced off.

%=============================================================================
\section{Panchromatic (SED) transport}\label{sec:sed}
%=============================================================================

\subsection{Grey rescaling}
In SED mode a packet is assigned a wavelength bin at birth and keeps it until
absorption.  Storing a separate opacity grid per wavelength would multiply the
memory by the number of bins; instead MoCafe stores the density structure --- and
hence the optical depth --- at a single reference wavelength $\lambda_{\rm ref}$
and rescales each packet by the wavelength-dependent opacity ratio
(Jonsson~2006),
\begin{equation}
  s_{\rm ext}(\lambda) = \frac{\kext(\lambda)}{\kext(\lambda_{\rm ref})},
  \qquad
  \tau(\lambda) = s_{\rm ext}(\lambda)\,\tau_{\rm ref}.
  \label{eq:grey}
\end{equation}
The ray tracers integrate $\tau_{\rm ref}$ and multiply by the packet's
$s_{\rm ext}$; the free-path draw of Eq.~\eqref{eq:freepath} is done in
$\tau_{\rm ref}=\tau(\lambda)/s_{\rm ext}$.  Because $s_{\rm ext}$ is a single
scalar per packet, all three tracers are shared with the monochromatic path and
the memory does not grow with the number of wavelengths.  The wavelength-dependent
albedo $a(\lambda)$ and asymmetry $g(\lambda)$ come from the same opacity table
($\lambda$, $a$, $\langle\cos\rangle$, $\Cext/$H).

\subsection{Wavelength sampling and packet luminosity}
The source spectrum $L_\lambda$ --- a Planck function at \pp{tstar}, a tabulated
\pp{source\_spectrum}, or a per-component spectrum in the multi-source case --- is
integrated over each bin to a luminosity fraction $f_i=L_i/L_{\rm tot}$, and the
birth bin is drawn from $\{f_i\}$ with an alias table.  Each packet then carries a
physical luminosity $L_{\rm p}=L_{\rm tot}/N_{\rm phot}$, so that the sum of
packet luminosities reproduces the source luminosity and the downstream emission
budget is absolute (erg\,s$^{-1}$).  Multiple stellar populations
(\S\ref{sec:galaxy}) are mixed by drawing the emitting component in proportion to
its luminosity.

%=============================================================================
\section{The radiation field: Lucy path-length estimator}\label{sec:jlam}
%=============================================================================

\subsection{Mean intensity}
Dust heating is set by the mean intensity (zeroth angular moment of $I_\lambda$)
in each cell,
\begin{equation}
  \Jlam(\rvec)=\frac{1}{4\pi}\!\int I_\lambda(\rvec,\nhat)\,\dd\Omega .
\end{equation}
MoCafe estimates $\Jlam$ with the Lucy~(1999) \emph{path-length} estimator.  The
energy density is $u_\lambda = (4\pi/c)\Jlam$, and the total path length that
radiation traverses in a volume $V$ per unit time equals $c$ times the number of
photons in $V$; accumulating a contribution from every segment a packet takes
through the cell therefore estimates $\Jlam$ directly:
\begin{equation}
  \Jlam = \frac{1}{4\pi\,V\,\Delta\lambda\,d_{\rm cm}^2}
          \sum_{\text{segments in cell}} L_{\rm p}\, w\, \ell,
  \label{eq:jlucy}
\end{equation}
where $\ell$ is the segment length, $V$ the cell volume, $\Delta\lambda$ the bin
width, and $d_{\rm cm}$ the length unit in cm.

\subsection{Path-length versus collision estimator}
A collision (or absorption) estimator would tally only at discrete interaction
points, so a cell with few interactions receives a noisy, or zero, estimate.
The path-length estimator instead credits every fly-through, whether or not the
packet interacts, so even optically thin cells --- where interactions are rare
but many packets pass through --- get a smooth, low-variance $\Jlam$.  This is
decisive for dust heating, where the illuminating field must be known in cells
that are optically thin to the starlight.

\subsection{Absorbed power and the analytic first flight}
The power absorbed in a cell follows from $\Jlam$ and the absorption opacity,
\begin{equation}
  P_{\rm abs} = 4\pi V \!\int \kabs(\lambda)\,\Jlam \, \dd\lambda .
  \label{eq:pabs}
\end{equation}
The forced first flight (\S\ref{sec:sed}) has a biased free-path distribution, so
that segment is \emph{not} tallied by sampling; instead its exact expectation is
added analytically along the ray to the edge,
\begin{equation}
  \Jlam \mathrel{+}= \frac{w\,L_{\rm p}}{4\pi V\Delta\lambda\,d_{\rm cm}^2}
     \int e^{-\tau_\lambda(\ell)}\,\dd\ell
   = \frac{w\,L_{\rm p}}{\dots}\,
     \frac{e^{-s\,\tau_{\rm in}}-e^{-s\,\tau_{\rm out}}}{\kext s},
\end{equation}
a zero-variance direct component.  Flights with $n_{\rm scatt}\ge1$ sample the
ordinary exponential free path and are tallied along the walked segments.  The
$\code{\_jlam}$ output reports the pathlength-tallied absorbed power
(\code{EABS\_A}) against an independent event counter (\code{EABS\_B}) as an
energy-conservation check.

%=============================================================================
\section{Dust thermal emission I: Lucy + SEDust}\label{sec:lucy}
%=============================================================================

The radiation field $\Jlam$ from Eq.~\eqref{eq:jlucy} is handed cell by cell to
the \textsc{SEDust} grain library, which returns the emission spectrum for the
local field.  Below, $\Cabs(a,\lambda)=\pi a^2 Q_{\rm abs}(a,\lambda)$ is the
absorption cross section of a grain of radius $a$.

\subsection{Grain energy balance}
A grain absorbs starlight at the rate
$\dot E_{\rm abs}=\int 4\pi\,\Cabs(a,\lambda)\,\Jlam\,\dd\lambda$ and radiates at
$\dot E_{\rm emit}(T)=\int 4\pi\,\Cabs(a,\lambda)\,B_\lambda(T)\,\dd\lambda$.  A
\emph{large} grain reaches a steady temperature $T_{\rm eq}$ where
$\dot E_{\rm emit}(T_{\rm eq})=\dot E_{\rm abs}$; because the absorbed energy per
photon is tiny compared with the grain heat capacity, its temperature barely
fluctuates.  A \emph{small} grain (or a PAH molecule), by contrast, has so small
a heat capacity that a single ultraviolet photon can raise its temperature by
hundreds of kelvin, after which it cools before the next hit.  Its temperature
is therefore a random variable, and the emission must be integrated over its
probability distribution.

\subsection{Enthalpy and heat capacity}
The internal energy (enthalpy) $U(T)$ of a grain is obtained from the Debye
model of the lattice vibrations (Draine \& Li 2001): silicate and graphite use
two- and three-dimensional Debye integrals with the appropriate Debye
temperatures, and PAHs add discrete C--H and C--C mode contributions
(\file{enthalpy\_v2.f90}, \file{enthalpy\_astrodust.f90}).  Inverting $U(T)$
gives the temperature reached after absorbing a given energy, which is the input
to the stochastic-heating solver.

\subsection{Stochastic heating: the temperature probability distribution}
\textsc{SEDust} solves for the steady-state probability distribution $P_j$ over
a set of discrete enthalpy bins $\{U_j\}$ ($j=1\ldots N$, default $N=200$) using
Draine's thermal-discrete method (\file{stoch\_qm.f90}; Guhathakurta \&
Draine~1989; Draine \& Li~2001).  Transitions between bins are collected into a
rate matrix $A_{fi}$ (the rate of going from bin $i$ to bin $f$):
\begin{itemize}
\item \textbf{Heating} ($f>i$): absorption of a photon of energy
      $U_f-U_i=hc/\lambda$ carries the grain upward, at rate
      \begin{equation}
        A_{fi} = \frac{4\pi\,\Jlam}{hc/\lambda}\,\Cabs(a,\lambda)\,
                 \frac{\dd\lambda}{\dd U_f},
      \end{equation}
      i.e.\ proportional to the photon arrival rate at the wavelength that
      bridges the two bins.
\item \textbf{Cooling} ($f<i$): thermal emission removes energy.  In the
      thermal-discrete treatment (\code{dbdis}, the production setting) all
      downward transitions are retained, with rates set by the Planck emission
      at the temperature of the upper bin; the continuous-cooling approximation
      (\code{dbcon}) collapses these into nearest-neighbour ($f=i-1$)
      transitions, recovering the Guhathakurta--Draine temperature-space
      recursion.
\end{itemize}
Conservation of probability fixes the diagonal, and the steady state satisfies
\begin{equation}
  \sum_{i\neq j} A_{ji}\,P_i = P_j \sum_{i\neq j} A_{ij},
  \qquad \sum_j P_j = 1,
\end{equation}
solved as a sparse linear system with a bi-conjugate-gradient solver.  The
result $P_j$ is the probability that the grain sits in enthalpy bin $j$, i.e.\
at temperature $T_j=T(U_j)$.

\subsection{Equilibrium limit}
For large grains the distribution $P_j$ collapses to a narrow spike at
$T_{\rm eq}$, and the machinery reduces to the energy-balance root
$\dot E_{\rm emit}(T_{\rm eq})=\dot E_{\rm abs}$.  \textsc{SEDust} detects this
regime and can use the equilibrium temperature directly, which is exact for the
big grains that dominate the far-infrared and much cheaper than the full solve.

\subsection{Size distribution and PAHs}
The cell emissivity integrates the single-grain emission over the grain-size
distribution $\dd n/\dd a$ of each material (silicate, carbonaceous, PAH),
\begin{equation}
  \varepsilon_\lambda = \sum_{\rm species}\int \dd a\;\frac{\dd n}{\dd a}\;
     4\pi\,\Cabs(a,\lambda)\int \dd T\; P(a,T)\, B_\lambda(T),
  \label{eq:emis}
\end{equation}
where $P(a,T)$ is the stochastic distribution above (a delta function at
$T_{\rm eq}$ for large grains).  The PAH population produces the mid-infrared
band spectrum (3.3, 6.2, 7.7, 8.6, 11.3\,$\mu$m); the smallest grains produce
the smoothly rising continuum from single-photon heating.  The size
distributions and optical constants come from the selected grain model.

\subsection{Emission spectrum and energy normalization}
The emission in each cell is renormalized so that its wavelength integral equals
the locally absorbed power of Eq.~\eqref{eq:pabs},
\begin{equation}
  \int \varepsilon_\lambda^{\rm norm}\,\dd\lambda = P_{\rm abs},
\end{equation}
which makes energy conservation \emph{exact} regardless of the internal
\textsc{SEDust} normalization.  The normalized spectrum is resampled onto the
transport wavelength grid and either integrated directly (the intrinsic dust
SED) or emitted as a second batch of packets --- sampled in wavelength from
$\varepsilon_\lambda^{\rm norm}$ and in position from the cell emissivity --- to
form the emergent SED and the pixel-resolved emission image.

\subsection{Speed options}
The exact stochastic solve costs $\mathcal O(0.1\,\mathrm{s})$ per cell.  Two
approximations trade accuracy for speed:
\begin{description}
\item[\pp{dust\_single\_teq}] replaces $P(a,T)$ by a single mixture equilibrium
  temperature.  It keeps the total power but drops the PAH and stochastic
  features (no mid-infrared bands), and is intended for optically thin,
  far-infrared-dominated problems.
\item[\pp{dust\_fast\_table}] exploits that, to good approximation, the emission
  spectrum depends on the field only through its intensity
  $U=\int\Jlam\,\dd\lambda/\int J_\lambda^{\rm ref}\,\dd\lambda$ and shape.  A
  reference emission table is built over a grid of $U$ once, and each cell is
  interpolated in $\log U$.  This preserves the PAH/stochastic features at
  $\sim1\%$ SED-shape error, with the largest deviations on the far-infrared
  slope.
\end{description}

\subsection{Self-absorption iteration}
At high infrared optical depth the dust re-absorbs its own emission.  With
\pp{dust\_niter}$>1$ MoCafe transports the emitted packets, tallies their
contribution to the field, adds it to the stellar field
$\Jlam\to\Jlam^{\star}+\Jlam^{\rm dust}$, and recomputes the emission, iterating
until the total emitted luminosity converges to \pp{dust\_tol}.  Writing $f$ for
the fraction of dust emission that is re-absorbed, the emitted luminosity obeys
\begin{equation}
  L^{(k)} = L_{\rm abs}^{\star} + f\,L^{(k-1)}
  \;\longrightarrow\;
  \frac{L_{\rm abs}^{\star}}{1-f},
\end{equation}
a geometric series.  At $\tau_V=100$ this converges in about ten iterations, and
only the converged run conserves energy: a single pass loses the re-absorbed
dust emission (only $\sim37\%$ of the starlight escapes), whereas the converged
run returns $99.8\%$.  Iteration is therefore \emph{required} for energy
conservation at high infrared optical depth.

%=============================================================================
\section{Dust thermal emission II: Bjorkman \& Wood}\label{sec:bw}
%=============================================================================

The Bjorkman \& Wood~(2001) scheme uses fixed-energy packets and a mixture mean
opacity, and needs no radiation-field storage and no iteration.  When a packet is
absorbed in a cell, the cell's cumulative absorbed energy increases, raising its
equilibrium temperature from $T$ to $T+\Delta T$; the packet is re-emitted
\emph{immediately} with a wavelength drawn from the temperature-correction
spectrum.  The construction is as follows.  A cell that has already emitted a
spectrum $\propto\kabs B_\lambda(T)$ must, after the temperature increment,
radiate $\propto\kabs B_\lambda(T+\Delta T)$.  The \emph{additional} photons
needed to correct the spectrum are therefore distributed as the difference,
\begin{equation}
  \frac{\dd P}{\dd\lambda}\;\propto\;
     \kabs(\lambda)\,\big[B_\lambda(T{+}\Delta T)-B_\lambda(T)\big]
  \;\approx\;
     \kabs(\lambda)\,\frac{\partial B_\lambda(T)}{\partial T},
  \label{eq:bwcorr}
\end{equation}
which is the sampling law for the reemitted packet.  Summed over all absorptions
this makes each cell radiate the correct $\kabs B_\lambda(T)$ spectrum in steady
state.  The equilibrium temperature is found by inverting the monotonic
$\kappa_{\rm P}(T)=\int\kabs(\lambda)\,B_\lambda(T)\,\dd\lambda$ against the
running absorbed power.  The method is approximate (equilibrium only, mixture
opacity, no PAH/stochastic features) but cheap, and the emergent dust light
lands directly in the observer \code{Scattered} SED.  Over $\tau_V=0.1\ldots20$
its absorbed luminosity agrees with the Lucy method to $<2\%$.

%=============================================================================
\section{Modified Random Walk}\label{sec:mrw}
%=============================================================================

In cells that are very optically thick to \emph{scattering}, a packet would take
an enormous number of tiny steps.  The Modified Random Walk (Min et al.~2009;
Robitaille~2010) replaces a whole diffusive excursion with one step.  A sphere of
radius $R_0$ (the distance to the nearest cell wall) is inscribed at the packet
position; MRW triggers when its scattering optical thickness exceeds a threshold,
$\ksca R_0 > \gamma$ (\pp{mrw\_gamma}, default 2).  The dimensionless first-passage
variable $y$ --- the probability that the diffusion has not yet reached the sphere
surface --- has cumulative distribution
\begin{equation}
  P(y) = 2\sum_{n=1}^{\infty}(-1)^{n+1}\,y^{\,n^2},\qquad y\in[0,1],
\end{equation}
which MoCafe tabulates and inverts once at setup.  A draw of $y$ fixes the total
\emph{scattering} path length travelled inside the sphere,
\begin{equation}
  ct = -\ln y \;\cdot\; 3\,\ksca \left(\frac{R_0}{\pi}\right)^{\!2},
\end{equation}
after which the packet is moved to a random point on the sphere with a fresh
isotropic direction.  MRW diffuses with the \emph{scattering} opacity; the
absorption along the path is applied as a weight decay
$w\leftarrow w\,e^{-\kabs\,ct}$, and the corresponding absorbed energy is
deposited into the radiation-field tally as the path integral
$\int_0^{ct} w\,L_{\rm p}\,e^{-\kabs\ell}\,\dd\ell$.  Russian roulette
terminates high-albedo, high-$\tau$ excursions.  Because realistic dust is
genuinely absorbing, the diffusion is usually terminated by absorption before
MRW helps; its speed-up is large only for near-pure-scattering, very thick cells,
so MRW is off by default and adds no measurable overhead when it rarely triggers.

%=============================================================================
\section{Source (galaxy) models}\label{sec:galaxy}
%=============================================================================

A stellar component emits packets from a spatial distribution; multiple
components (\pp{nsource}) are selected in proportion to luminosity, and each may
have its own spectrum and geometry (bulge $+$ disk gives the bulge-to-total
ratio).

\subsection{Exponential disk}
A razor-thin exponential disk has surface density
$\Sigma(r)\propto e^{-r/r_s}$, so the probability of emitting from cylindrical
radius $r$ is
\begin{equation}
  p(r)\,\dd r \propto 2\pi r\,\Sigma(r)\,\dd r \propto r\,e^{-r/r_s}\,\dd r,
\end{equation}
the $\mathrm{Gamma}(2)$ (double-exponential) law.  It is sampled \emph{exactly}
as the sum of two exponential deviates, $r=-r_s(\ln\xi_1+\ln\xi_2)$
(\code{rand\_r1exp}), avoiding any inverse-CDF search.  The azimuth is uniform
(modified by spiral arms, below).

\subsection{Vertical structure}
The vertical profile is exponential, $\rho\propto e^{-|z|/z_s}$
(\code{rand\_zexp}), or the self-gravitating isothermal-sheet form
$\rho\propto\mathrm{sech}^2(z/z_s)$ (\code{rand\_sech2}), selected by
\pp{density\_zprofile}.  The same options apply to the dust disk, whose density
is the double exponential $\exp(-|z|/z_s)\exp(-r/r_s)$.

\subsection{Spiral arms}
Log-spiral arms multiply the disk by
\begin{equation}
  1 + A\,\sin\!\Big[m\Big(\frac{\ln r}{\tan p}-\phi\Big)\Big],
\end{equation}
with $m$ arms, pitch angle $p$, and amplitude $A$ (\pp{spiral\_m},
\pp{spiral\_pitch}, \pp{spiral\_amp}).  For \emph{sources} the azimuth $\phi$ is
drawn by rejection against this factor (accept if
$(1+A)\,\xi\le 1+A\sin[\dots]$); for the \emph{dust} density the cell opacity is
multiplied by the same factor.  Seen face-on, the absorbed-power (or
dust-emission) map traces the arms, validated in
\file{examples/galaxy/spiral\_faceon.in}.

\subsection{Sersic bulge: deprojection and alias sampling}\label{sec:sersic}
The projected Sersic surface-brightness profile is
$I(R)=I_0\exp[-b_m(R/R_e)^{1/m}]$ with index $m$ (\pp{src\_sersic\_index}),
effective radius $R_e$ (\pp{src\_reff}), and $b_m$ from Ciotti \& Bertin~(1999)
so that $R_e$ encloses half the projected light.  MoCafe emits from the
\emph{three-dimensional} luminosity density $\nu(r)$, recovered from $I(R)$ by the
inverse Abel transform,
\begin{equation}
  \nu(r) = -\frac{1}{\pi}\int_r^\infty \frac{\dd I}{\dd R}\,
             \frac{\dd R}{\sqrt{R^2-r^2}} .
\end{equation}
The deprojected cumulative luminosity within spherical radius $r$ is
\begin{equation}
  \mathcal L(<r) = \gamma\!\big(2m{+}1,\,b_m r^{1/m}\big)
     + \frac{2}{\pi}\frac{b_m^{\,2m+1}}{m\,\Gamma(2m{+}1)}\,r^2
       \!\int_1^\infty\!\! x\,f(x)\,e^{-b_m(rx)^{1/m}}(rx)^{1/m}\,\dd x,
\end{equation}
computed once on a logarithmic radius grid (\code{sersic\_cumulative\_3D}, with
the incomplete gamma function $\gamma$).  The radius is then drawn with the
\emph{alias} method (\S\ref{sec:alias}) rather than an inverse-CDF search: the
bin probabilities (differences of the cumulative) build an alias table, and
\code{rand\_alias\_linear} returns a continuous radius by linear interpolation
within the selected bin.  The point is placed isotropically and flattened in $z$
by the axial ratio (\pp{src\_axial\_ratio}).  The alias-sampled distribution
reproduces the analytic cumulative to $\sim10^{-4}$.

\subsection{Boxy / bar / x-bar bulges}
Generalized-ellipsoid bulges use the generalized radius
$r=(|x|^n+|y|^n+|z|^n)^{1/n}$ with boxiness $n$ (\pp{src\_boxiness}; $n=2$ is an
ellipsoid, $n>2$ boxy) and a power-law (\code{boxy}), Gaussian (\code{bar}), or
peanut (\code{xbar}) radial profile.  These are drawn with a slice-within-Gibbs
sampler (\file{random\_bulge.f90}): an auxiliary height is drawn uniformly under
the current density value, and each coordinate is then updated uniformly on the
resulting horizontal slice.  Consecutive draws are correlated (the sampler
explores by conditional updates), which is adequate for a smooth bulge but
statistically weaker than the independent alias-sampled Sersic.

%=============================================================================
\section{Random sampling}\label{sec:alias}
%=============================================================================

\subsection{Inversion and rejection}
Two elementary methods recur above.  \emph{Inversion} solves $P(x)=\xi$ for the
cumulative $P$; it is exact and consumes one deviate, used for the free path
(Eq.~\ref{eq:freepath}), the Henyey--Greenstein cosine, and the exponential
disk.  \emph{Rejection} proposes $x$ from an envelope and accepts with
probability proportional to the target density; it needs no cumulative and is
used for the spiral azimuth and inside the Gibbs bulge sampler, at the cost of
occasional rejected proposals.

\subsection{Alias method}
Walker's alias method samples an index from a discrete distribution $\{p_k\}$
($k=1\ldots K$) in $\mathcal O(1)$ time after an $\mathcal O(K)$ setup.  The
setup (\code{random\_alias\_setup}, Vose's construction) rewrites the
distribution as a set of $K$ equal-probability ``columns,'' each holding at most
two outcomes: it stores an acceptance probability $q_k$ and an alias $A_k$ for
every bin.  A draw picks a uniform bin $k$ and a uniform $\eta\in[0,1)$, and
returns $k$ if $\eta<q_k$, else $A_k$.  MoCafe uses it for wavelength sampling
(\S\ref{sec:sed}), the Mueller $S_{11}$ angle, and the Sersic radius --- anywhere
a fixed discrete distribution is sampled many times.

\subsection{Continuous piecewise-linear alias}
For a continuous variable tabulated at nodes $x_i$ with density $p_i$, the
probability of the interval $[x_i,x_{i+1}]$ is the trapezoid area
$\tfrac12(p_i+p_{i+1})(x_{i+1}-x_i)$.  An alias table over these interval
probabilities selects a bin in $\mathcal O(1)$; within the bin the value is
returned by inverting the linear density (\code{rand\_alias\_linear}), giving a
continuous, independent draw.  This is used for the Sersic radius and for
external-illumination angles.

%=============================================================================
\section{HEALPix all-sky interior observer}\label{sec:allsky}
%=============================================================================

For an observer \emph{inside} the medium (the Milky-Way case) there is no single
image plane.  Each scattering, dust-reemission, and direct event peels toward the
observer point, and the expected contribution (Eq.~\ref{eq:peel}, with the
$1/d_o^2$ replaced by the pixel solid angle) is binned into the HEALPix pixel of
the arrival direction $\nhat$.  HEALPix (G\'orski et al.~2005) tiles the sphere
into $n_{\rm pix}=12\,n_{\rm side}^2$ equal-area pixels of solid angle
$\Omega=4\pi/n_{\rm pix}$; the RING scheme is used, and the pixel index is found
by \code{vec2pix}.  The optical depth to the observer is integrated along the
matching ray (Cartesian or AMR).  The output carries the standard HEALPix
keywords so it reads directly with \code{healpy}.  Because each event maps to the
single pixel of its \emph{source} direction, a diffuse map needs many events
distributed over the sky; concentrated (inner-disk) far-infrared emission is
sparse at high $n_{\rm side}$ and benefits from a finer spatial grid or
smoothing.

%=============================================================================
\section{Energy conservation and validation}
%=============================================================================

Energy conservation is the main internal check and holds by construction:
\begin{itemize}
\item The total stellar luminosity equals the escaped starlight plus the
      absorbed (hence re-emitted) power.
\item The dust solvers normalize the emission to the absorbed power cell by
      cell, so the integrated intrinsic dust luminosity equals \code{L\_ABS} to
      machine precision, and the emergent SED integrated over $4\pi d^2$ recovers
      the same value to Monte-Carlo accuracy.
\item The $\code{\_jlam}$ output cross-checks the path-length-tallied absorbed
      power against an independent event counter.
\end{itemize}
Against external references, the \textsc{SEDust} single-cell emission tracks the
Camps et al.~(2015) seven-code stochastic-heating benchmark (which includes
\textsc{SKIRT} and \textsc{DIRTY}) to a 2--4\% median relative difference over
$U=10^{-2}\ldots10^{4}$, the Lucy and Bjorkman \& Wood absorbed luminosities
agree to $<2\%$ over $\tau_V=0.1\ldots20$, and dust emission on a uniform octree
sphere reproduces the equivalent Cartesian sphere.  These are consolidated in the
\file{examples/benchmarks/validate.py} regression harness.

%=============================================================================
\section{Summary of key numerical parameters}
%=============================================================================
\begin{table}[htb]
\centering
\caption{Parameters that control the algorithms in this note.}
\begin{tabular}{lll}
\toprule
Parameter & Meaning & Default \\
\midrule
\pp{use\_reduced\_wgt}  & forced first scattering + reduced weight & \code{.true.}\\
\pp{use\_stokes}        & polarized Mueller-matrix scattering      & \code{.false.}\\
\pp{lambda\_ref}        & reference wavelength for grey rescaling  & --- \\
\pp{dust\_model\_sed}   & grain model (\code{astrodust}/\code{dl07}/\code{zubko}) & \code{astrodust}\\
\pp{dust\_single\_teq}  & equilibrium-only emission (fast)         & \code{.false.}\\
\pp{dust\_fast\_table}  & $U$-interpolated emission table          & \code{.false.}\\
\pp{dust\_niter}        & self-absorption iterations               & 1 \\
\pp{dust\_tol}          & iteration convergence tolerance          & --- \\
\pp{mrw\_gamma}         & MRW trigger ($\ksca R_0>\gamma$)         & 2 \\
\pp{nside}              & HEALPix resolution (interior observer)   & --- \\
\bottomrule
\end{tabular}
\end{table}

%=============================================================================
\section*{References}
\addcontentsline{toc}{section}{References}
%=============================================================================
\begin{itemize}\small
\item Lucy, L.~B. 1999, A\&A, 344, 282 --- path-length mean-intensity estimator.
\item Bjorkman, J.~E. \& Wood, K. 2001, ApJ, 554, 615 --- immediate reemission
      with temperature correction.
\item Guhathakurta, P. \& Draine, B.~T. 1989, ApJ, 345, 230 --- stochastic
      heating, temperature-space recursion.
\item Draine, B.~T. \& Li, A. 2001, ApJ, 551, 807 --- thermal-discrete stochastic
      heating; grain enthalpies.
\item Draine, B.~T. \& Hensley, B.~S. 2021, ApJ, 909, 94 --- astrodust model.
\item Camps, P.\ et al. 2015, A\&A, 580, A87 --- stochastic-heating dust-emission
      benchmark.
\item Draine, B.~T. \& Li, A. 2007, ApJ, 657, 810 --- DL07 dust model.
\item Zubko, V., Dwek, E., \& Arendt, R.~G. 2004, ApJS, 152, 211 --- BARE-GR-S
      model.
\item Min, M.\ et al. 2009, A\&A, 497, 155; Robitaille, T.~P. 2010, A\&A, 520,
      A70 --- Modified Random Walk.
\item Jonsson, P. 2006, MNRAS, 372, 2 --- grey rescaling / polychromatic scan.
\item Ciotti, L. \& Bertin, G. 1999, A\&A, 352, 447 --- Sersic $b_m$.
\item Walker, A.~J. 1977, ACM TOMS, 3, 253; Vose, M.~D. 1991, IEEE TSE, 17, 972
      --- the alias method.
\item G\'orski, K.~M.\ et al. 2005, ApJ, 622, 759 --- HEALPix pixelization.
\item Seon, K.-I. 2010, PKAS, 25, 177 --- the $(a,g)$ single-run scan.
\item Amanatides, J. \& Woo, A. 1987, Eurographics~'87 --- fast voxel traversal.
\end{itemize}

\end{document}
